Non-word repetition in Yélî Dnye

### 2021-02-10 manuscript.Rmd
minor wordings (incl. new ethngphy bit discussed via email)

Yélî Dnye is an isolate language (presumed Papuan) spoken by approximately 7,000 people residing on Rossel Island, an island found at the far end of the Louisiade Archipelago in Milne Bay Province, Papua New Guinea. The Yélî sound system, much like its baroque grammatical system [@levinsonYDgrammar], is unlike any other in the region. In total, Yélî Dnye uses 90 distinctive segments (not including an additional three rarely used consonants), far outstripping the phonemic inventory size of other documented Papuan languages [@foley1986papuan; @levinsonYDgrammar; @maddiesonIPphoneticsYD]. Thus, with respect to our first research goal, Yélî Dnye is a good language to attempt an investigation on NWR with sounds varying in cross-linguistic frequency because of its large inventory, which includes some rare sounds.

@@

Most Yélî parents are swidden horticulturalists. Within a group of households, it is often the case that older adolescent and adults spend their day tending to their gardens (which may not be nearby), bringing up water from the river, washing clothes, preparing food, and engaging in other such activities, which leaves them less time to spend directly with the children in their household than may be the case for stay-at-home parents in urban sites. Starting around age two years, children more often spend large swaths of their day playing, swimming, and foraging for fruit, nuts, and shellfish in large (~10 members) independent and mixed-age child play groups [@brownIPchildrearing; @casillas2020early]. Formal education is a priority for Yélî families, and many young parents have themselves pursued additional education beyond of what is locally available [@casillas2020early]. Local schools are well out of walking distance for many children (i.e., more than 1 hour on foot or by canoe each day), so it is very common for households situated close to a school to billet their school-aged relatives during the weekdays for long segments of the school year. Children start school often at around age seven, although the precise age depends on the child's readiness, as judged by their teacher.

@@

In a basic NWR task, the participant listens to a production of a word-like form, such as /bilik/, and then repeats back what they heard without changing any phonological feature that is contrastive in the language. For instance, in English, a response of [bilig] or [pilik] would be scored as incorrect; a response [biːlik], where the vowel is lengthened without change of quality would be scored as correct, because English does not have contrastive vowel length. There is some variation in how past NWR studies have designed the presentation procedure and structure of items. For example, while items are often presented orally by the experimenter [@torrington2015non], an increasing number of studies have turned instead to playing back pre-recorded stimuli in order to increase control in stimulus presentation [@brandeker2015language]. Additionally, while some studies have used 10-15 non-words [e.g., @cristia2020infant], others have employed up to 46 unique items [@piazzalunga2019articulatory]. Authors also often modulate structural complexity, typically measured in terms of item length (measured in number of syllables) and/or syllable structure [open as opposed to closed syllables, @gallon2007non].

@@

We fitted the child with a headset microphone (Shure SM10A or WH20 XLR with a dynamic microphone on a headband, most children using the former) that fed into the left channel of a Tascam DR40x digital audio recorder. The headsets were designed for adult use and could not be comfortably seated on many children's heads without a more involved adjustment period. To minimize adjustment time, which was uncomfortable for some children given the proximity of the experimenter and equipment, we placed the headband on children's shoulders in these cases, carefully adjusting the microphone's placement so that it was still close to the child's mouth. A research assistant who spoke Yélî Dnye natively sat next to the child throughout the task to provide instructions and, if needed, encouragement. The research assistant coached the child throughout the task to make sure that they understood what they were expected to do. An experimenter (the first author) delivered the pre-recorded stimuli to the research assistant and the child over headphones.

@@

We tested a total of `r dim(demo)[1]` children from `r length(levels(factor(demo$familyID)))` families spread across four hamlet regions. We excluded test sessions from analysis for the following reasons: refused participation or failure to repeat items presented over headphones even after coaching (N=`r sum(demo$included=="didn\xeat repeat")`), spoke too softly to allow offline coding (N=`r sum(demo$included=="spoke too softly")`), or were 13 years old or older (N=`r sum(demo$Age >12 | demo$included=="too old",na.rm=T)`; we tested these teenagers to put younger children at ease). The remaining `r dim(inc)[1]` children (`r sum(inc$Sex=="G")` girls) were aged from `r min(inc$Age, na.rm=T)` to `r max(inc$Age, na.rm=T)` years (M = `r round(mean(inc$Age, na.rm=T),1)` years, SD = `r round(sd(inc$Age, na.rm=T),1)` years).  In terms of birth order, `r birthorder.tab[[1]]` were first borns, `r birthorder.tab[[2]]` second,  `r birthorder.tab[[3]]` third,  `r birthorder.tab[[4]]` fourth,  `r birthorder.tab[[5]]` fifth, and `r birthorder.tab[[6]]` sixth, with birth order missing for `r sum(is.na(inc$birthOrder))` children. These children were tested in a hamlet far from our research base, and we unfortunately did not ask about birth order before leaving the site. Maternal years of education averaged `r round(mean(inc$matEd,na.rm=T),2)` years (range `r round(min(inc$matEd,na.rm=T),2)`-`r round(max(inc$matEd,na.rm=T),2)` years).[^3] We also note that there were `r sum(inc$monolingual=="yes", na.rm=T)` only exposed to Yélî Dnye at home and `r sum(inc$monolingual=="no", na.rm=T)` children exposed to Yélî Dnye plus one or more other languages at home.[^2]

>>>>

Yélî Dnye is an isolate language (presumed Papuan) spoken by approximately 7,000 people residing on Rossel Island, an island found at the far end of the Louisiade Archipelago in Milne Bay Province, Papua New Guinea. The Yélî sound system, much like its baroque grammatical system [@levinsonYDgrammar], is unlike any other in the region. In total, Yélî Dnye uses 90 distinctive segments (not including an additional three rarely used consonants), far outstripping the phonemic inventory size of other documented Papuan languages [@foley1986papuan; @levinsonYDgrammar; @maddiesonIPphoneticsYD]. Thus, with respect to our first research goal, Yélî Dnye is a good language to test because its large phonological inventory includes sounds that vary in cross-linguistic frequency (including some rare sounds) that can be compared in the NWR setting.

@@

Most Yélî parents are swidden horticulturalists. Within a group of households, it is often the case that older adolescents and adults spend their day tending to their gardens (which may not be nearby), bringing up water from the river, washing clothes, preparing food, and engaging in other such activities. Starting around age two years, children more often spend large swaths of their day playing, swimming, and foraging for fruit, nuts, and shellfish in large (~10 members) independent and mixed-age child play groups [@brownIPchildrearing; @casillas2020early]. Formal education is a priority for Yélî families, and many young parents have themselves pursued additional education beyond of what is locally available [@casillas2020early]. Local schools are well out of walking distance for many children (i.e., more than 1 hour on foot or by canoe each day), so it is very common for households situated close to a school to billet their school-aged relatives during the weekdays for long segments of the school year. Children start school often at around age seven, although the precise age depends on the child's readiness, as judged by their teacher.

@@

In a basic NWR task, the participant listens to a production of a word-like form, such as /bilik/, and then repeats back what they heard without changing any phonological feature that is contrastive in the language. For instance, in English, a response of [bilig] or [pilik] would be scored as incorrect; a response [biːlik], where the vowel is lengthened without change of quality would be scored as correct, because English does not have (unconditioned) contrastive vowel length. There is some variation in how past NWR studies have designed the presentation procedure and structure of items. For example, while items are often presented orally by the experimenter [@torrington2015non], an increasing number of studies have turned instead to playing back pre-recorded stimuli in order to increase control in stimulus presentation [@brandeker2015language]. Additionally, while some studies have used 10-15 non-words [e.g., @cristia2020infant], others have employed up to 46 unique items [@piazzalunga2019articulatory]. Authors also often modulate structural complexity, typically measured in terms of item length (measured in number of syllables) and/or syllable structure [open as opposed to closed syllables, @gallon2007non].

@@

We fitted the child with a headset microphone (Shure SM10A or WH20 XLR with a dynamic microphone on a headband, most children using the former) that fed into the left channel of a Tascam DR40x digital audio recorder. The headsets were designed for adult use and could not be comfortably seated on many children's heads without a more involved adjustment period. To minimize adjustment time, which was uncomfortable for some children given the proximity of the foreign experimenter and equipment, we placed the headband on children's shoulders in these cases, carefully adjusting the microphone's placement so that it was still close to the child's mouth. A research assistant who spoke Yélî Dnye natively sat next to the child throughout the task to provide instructions and, if needed, encouragement. The research assistant coached the child throughout the task to make sure that they understood what they were expected to do. An experimenter (the first author) delivered the pre-recorded stimuli to the research assistant and the child over headphones.

@@

We tested a total of `r dim(demo)[1]` children from `r length(levels(factor(demo$familyID)))` families spread across four hamlet regions. We excluded test sessions from analysis for the following reasons: refused participation or failure to repeat items presented over headphones even after coaching (N=`r sum(demo$included=="didn\xeat repeat")`), spoke too softly to allow offline coding (N=`r sum(demo$included=="spoke too softly")`), or were 13 years old or older (N=`r sum(demo$Age >12 | demo$included=="too old",na.rm=T)`; we tested these teenagers to put younger children at ease). The remaining `r dim(inc)[1]` children (`r sum(inc$Sex=="G")` girls) were aged from `r min(inc$Age, na.rm=T)` to `r max(inc$Age, na.rm=T)` years (M = `r round(mean(inc$Age, na.rm=T),1)` years, SD = `r round(sd(inc$Age, na.rm=T),1)` years).  In terms of birth order, `r birthorder.tab[[1]]` were first borns, `r birthorder.tab[[2]]` second,  `r birthorder.tab[[3]]` third,  `r birthorder.tab[[4]]` fourth,  `r birthorder.tab[[5]]` fifth, and `r birthorder.tab[[6]]` sixth, with birth order missing for `r sum(is.na(inc$birthOrder))` children. These children were tested in a hamlet far from our research base, and we unfortunately did not ask about birth order before leaving the site. Maternal years of education averaged `r round(mean(inc$matEd,na.rm=T),2)` years (range `r round(min(inc$matEd,na.rm=T),2)`-`r round(max(inc$matEd,na.rm=T),2)` years).[^3] We also note that there were `r sum(inc$monolingual=="yes", na.rm=T)` children only exposed to Yélî Dnye at home and `r sum(inc$monolingual=="no", na.rm=T)` children exposed to Yélî Dnye plus one or more other languages at home.[^2]
